Supplementary Report for the article *Functional diversity can facilitate the collapse of an undesirable ecosystem state*

### 2023-01-30 reports/supplement/supplement_0.9.1.Rmd
changes supplement

We have confirmed that when there is no within functional group strain variation, the dynamics of the model are not affected by the number of strains in each of the functional groups. This is a necessary condition for such a model [Moisset de Espanés et al. 2021](https://onlinelibrary.wiley.com/doi/10.1111/ele.13775), and gives confidence that our methods and their implementation are correct and robust.

@@

# Log and non-log oxygen diffusivity results

(ref:fig-log-nonlog) The effect of trait diversity (y-axis) viewed with a log axis (left panels) or arithmetic axis for the oxygen diffusivity gradient (x-axis). The positions of the tipping points (anoxic to oxic in blue, oxic to anoxic in red) and the extent of the region of bistability (green or orange shaded region) are shown. The left-most region is anoxic only, the middle is oxic or anoxic depending on historical conditions, and the right-most is oxic only. In the left panels the thick grey vertical lines at -8 and 0 on the x-axis show the extent of the oxygen diffusivity gradient investigated in the simulations. In panels e, f, i, and j the orange region shows when patterns of coexistence were atypical of the classical pattern shown in Figure 1a of the main text (e.g. when cyanobacteria coexisted with sulfur bacteria at low oxygen diffusivities). In the right-hand panels (arithmetic x-axis scale) the blue anoxic to oxic tipping point line often has steps. This is becase the simulated oxygen diffusivity levels were evenly distributed on the log-oxygen diffusivity scale, and so are relatively far apart at high oxygen diffusivities on the arithmetic scale.

@@

>>>>

We have confirmed that when there is no within functional group strain variation, the dynamics of the model are not affected by the number of strains in each of the functional groups. This is a necessary condition for such a model ([Moisset de Espanés et al. 2021](https://onlinelibrary.wiley.com/doi/10.1111/ele.13775)), and gives confidence that our methods and their implementation are correct and robust.

@@

# Log and non-log oxygen diffusivity

For comparability with the [Bush et al publication 2017](https://www.nature.com/articles/s41467-017-00912-x), we use a log-scale of the oxygen diffusivity gradient when displaying the effect of trait variation on tipping point positions. Throughout the text, we do not make comparisons when their outcome would depend on the decision to measure effect sizes on a log-scale of the environmental driver. For example, we compare the direction but not the magnitude of the diversity effects of the three functional groups on the resilience of the ecosystem state that they dominate (CB: oxic-anoxic tipping point; SB and PB: anoxic-oxic tipping point). On a logarithmic scale, the magnitude of these diversity effects depends on whether a tipping point is located at low or high oxygen diffusivity (Figure 13.1) and therefore on whether a group dominates that oxic or the anoxic state.

(ref:fig-log-nonlog) The effect of trait diversity (y-axis) viewed with a log axis (left panels) or arithmetic axis (right panels) for the oxygen diffusivity gradient (x-axis). The positions of the tipping points (anoxic to oxic in blue, oxic to anoxic in red) and the extent of the region of bistability (green or orange shaded region) are shown. The left-most region is anoxic only, the middle is oxic or anoxic depending on historical conditions, and the right-most is oxic only. In the left panels the thick grey vertical lines at -8 and 0 on the x-axis show the extent of the oxygen diffusivity gradient investigated in the simulations. In panels e, f, i, and j the orange region shows when patterns of coexistence were atypical of the classical pattern shown in Figure 1a of the main text (e.g. when cyanobacteria coexisted with sulfur bacteria at low oxygen diffusivities). In the right-hand panels (arithmetic x-axis scale) the blue anoxic to oxic tipping point line often has steps. This is because the simulated oxygen diffusivity levels were evenly distributed on the log-oxygen diffusivity scale, and so are relatively far apart at high oxygen diffusivities on the arithmetic scale.

@@

Moisset de Espanés, P., Ramos-Jiliberto, R. & Soto, J.A. (2021). Avoiding artifacts when varying the number of species in ecological models. Ecol. Lett., 24, 1976–1987.
